---
title: "Assignment 5"
author:
  - xw-zeng
date: "2022-12-14"
documentclass: ctexart
geometry: "left=3.18cm, right=3.18cm, top=2.54cm, bottom=2.54cm"
output:
  rticles::ctex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Problem sets.

- `9.1`
- `9.4 (a)`
- `9.6 (a)(b)(c)(d)`

List of functions.

- `Bootstrap_Res`: Bootstrap regression with the method of bootstrapping the residuals.
- `Bootstrap_Case`: Bootstrap regression with the method of bootstrapping paired cases.
- `Subsample_Bootstrap`: Use Subsampling + Bootstrapping method to determine the optimal block size.
- `Moving_Block`: Moving block Bootstrap method.
- `Model_Based_np`: Assume an AR(1) model and bootstrap this model the method of bootstrapping the residuals.
- `Model_Based_p`: Assume an AR(1) model s.t. errors following normal distribution and bootstrap this model the method of bootstrapping the errors.
- `Blocks_of_Blocks`: Moving block bootstrap method with Blocks-of-Blocks strategy.

Load the R packages.
```{r}
library(moments)
library(ggplot2)
```

\newpage

# 9.1

\vspace{-0.5cm}

$$
\begin{aligned}
&X_1, X_2, ..., X_n \sim \text{i.i.d. Bernoulli} (\theta) \\
&R(\mathcal X, F)=\bar X - \theta,\quad R^*=R(\mathcal X^*, \widehat F)=\bar {X^*} - \bar X\\
&\bar X=\frac{1}{n}\sum_{i=1}^{n}X_i, \quad \bar {X^*}=\frac{1}{n^*}\sum_{i=1}^{n^*}X^*_i\\
&X^*_1, X^*_2, ..., X^*_n \sim \text{i.i.d. Bernoulli} (\bar X) \sim \widehat F\\
&Let \ Z^*=\sum_{i=1}^{n^*}X^*_i \sim \text{i.i.d. Binomial} (n^*, \bar X)\\
&E^*(R^*)=E(\frac{1}{n^*}Z^*-\bar X)=\frac{1}{n^*}n^*\bar X-\bar X=0\\
&Var^*(R^*)=Var(\frac{1}{n^*}Z^*)=\frac{1}{{n^*}^2}n^*\bar X(1-\bar X)=\frac{1}{n^*}\bar X(1-\bar X)\\
\end{aligned}
$$

# 9.4

## (a)

The classic Beverton–Holt model for the relationship between spawners and recruits is:
$$
\begin{aligned}
&R=\frac{1}{\beta_{1}+\beta_{2} / S} \\
&\frac{1}{R}=\beta_1+\beta_2 \frac {1} {S}, \quad \beta_{1} \geq 0 \text { and } \beta_{2} \geq 0
\end{aligned}
$$

where $R$ and $S$ are the numbers of recruits and spawners, respectively.

Let $T=R+S$. $T$ is the total population abundance, which will only stabilize if $R=S$.
$$
\begin{aligned}
&\frac{2}{T}=\beta_1+\beta_2 \frac {2} {T} \\
&T=\frac{2(1-\beta_2)}{\beta_1} \Rightarrow R^*=S^*=T^*/2=\frac{1-\beta_2^*}{\beta_1^*}
\end{aligned}
$$

Load the data, and transform $R$, $S$ to new variables $\frac{1}{R}$,  $\frac{1}{S}$.
```{r}
data <- read.table('salmon.dat', header = TRUE)
data$R1 <- 1 / data$recruits; data$S1 <- 1 / data$spawners
```

First obtain the observed data estimate of $\hat R$.
```{r}
fit <- lm(R1 ~ S1, data)
R_hat <- (1 - fit$coefficients[2]) / fit$coefficients[1]; as.numeric(R_hat)
```

\newpage

### 1. Bootstrapping the residuals.

```{r}
Bootstrap_Res <- function(B){
  
  ###INITIAL VALUES###
  fit <- lm(R1 ~ S1, data)
  res <- residuals(fit)
  R_star <- rep(NA, B)
  
  ###MAIN###
  for (i in 1:B){
    res_bootstrap <- sample(res, length(res), replace = T)
    data$R2 <- fit$fitted.values + res_bootstrap
    fit2 <- lm(R2 ~ S1, data)
    beta_star <- fit2$coefficients
    R_star[i] <- (1 - beta_star[2]) / beta_star[1]
  }
  
  ###OUTPUT###
  hist(R_star, main = paste0('Histogram of Bootstrap (', B, ')'))
  return (R_star)
}
```

Histogram the bootstrap distribution.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result1 <- Bootstrap_Res(10000)
```

Obtain the corresponding 95% confidence interval.
```{r}
quantile(result1, c(0.025, 0.975))
```

Obtain the mean and standard error for $R^*$.
```{r, results='hold'}
print(paste('Mean:', mean(result1)))
print(paste('Standard Error:', sd(result1)))
```

Obtain the skewness and kurtosis for $R^*$.
```{r, results='hold'}
print(paste('Skewness:', skewness(result1)))
print(paste('Kurtosis:', kurtosis(result1)))
```

### 2. Bootstrapping the cases.
```{r}
Bootstrap_Case <- function(B){
  
  ###INITIAL VALUES###
  R1 <- data$R1; S1 <- data$S1; R_star <- rep(NA, B)
  
  ###MAIN###
  for (i in 1:B){
    idx <- sample(1:length(R1), length(R1), replace = T)
    R2 <- R1[idx]; S2 <- S1[idx]
    fit2 <- lm(R2 ~ S2)
    beta_star <- fit2$coefficients
    R_star[i] <- (1 - beta_star[2]) / beta_star[1]
  }
  
  ###OUTPUT###
  hist(R_star, main = paste0('Histogram of Bootstrap (', B, ')'))
  return (R_star)}
```

Histogram the bootstrap distribution.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result2 <- Bootstrap_Case(10000)
```

Obtain the corresponding 95% confidence interval.
```{r}
quantile(result2, c(0.025, 0.975))
```

Obtain the mean and standard error for $R^*$.
```{r, results='hold'}
print(paste('Mean:', mean(result2)))
print(paste('Standard Error:', sd(result2)))
```

Obtain the skewness and kurtosis for $R^*$.
```{r, results='hold'}
print(paste('Skewness:', skewness(result2)))
print(paste('Kurtosis:', kurtosis(result2)))
```

### 3. Compare the results of two methods.

`BR` is referred to the method of Bootstrapping the residuals, and similarly `BC` is referred to the method of Bootstrapping the cases.

1.\ Bootstrap distribution:

- The bootstrap distributions of both methods are nearly symmetric, but there exist some outliers on the left side of the distribution of `BC`.
- Kurtosis: `BR` < `BC`

2.\ Mean, standard error and confidence interval:

- Mean: `BR` > `BC`
- Standard Error: `BR` > `BC`
- Length of 95% Confidence Interval: `BR` > `BC`

# 9.6

Load the data and difference the data.
```{r}
data <- read.table('earthquake.dat', header = TRUE)
data <- diff(data$quakes, lag = 1)
```

## (a)

Use `Subsampling + Bootstrapping` method to find the best block size.

**Step 1** Find subsets of smaller size $m<n$, say $\mathcal{X}_i^{(m)}=(X_i,...,X_{i+m-1})$.

**Step 2** Bootstrap counterparts applied in $\mathcal{X}_i^{(m)}$, for a fixed $l^\prime$, $\hat{\phi}_{i, m}\left(l^{\prime}\right)=B^{-1} \sum_{b=1}^{B} H\left(\bar{X}_{i, b}^{*}\right)-H\left(\bar{X}_{i}\right)$.

**Step 3** Estimate $\widehat{\phi}_{m}(l^{\prime})$ based on $\widehat{\mathrm{MSE}}\left\{\widehat{\phi}_{m}\left(l^{\prime}\right)\right\}=\frac{1}{n-m+1} \sum_{i=1}^{n-m+1}\left(\hat{\phi}_{i, m}\left(l^{\prime}\right)-\hat{\phi}_{0}\left(l_{0}\right)\right)^{2}$.

**Step 4** Find $l_{\text {opt}}^{(m)^{\prime}}=\arg \min \widehat{\mathrm{MSE}}\left\{\widehat{\phi}_{m}\left(l^{\prime}\right)\right\}$, then $l_{\text {opt}}^{(m)^{\prime}} \approx\left(\frac{2 c_{2}^{2}}{c_{1}}\right)^{1/3} m^{1/3}$.

**Step 5** Rescale and find $l_{\text {opt}} \approx l_{\text {opt}}^{(m)^{\prime}}(n/m)^{1/3}$.

In this problem, we don't exactly know what parameter we need to estimate and in general the parameter to be estimated is usually a function of $\mu$ with $\mu$ as the mean, so to find $l_{\text {opt}}$ we can simply suppose that $\theta=\mu$.

Define the function of `Subsampling + Bootstrapping` method.
```{r}
Subsample_Bootstrap <- function(df, m, l0, B, l){
  
  ###INITIAL VALUES###
  n <- length(df)
  subsets <- matrix(NA, nrow = n - m + 1, ncol = m)
  for (i in 1:(n - m + 1)){subsets[i, ] <- df[i:(i + m - 1)]}
  
  ###FUNCTIONS###
  generate_phi_i <- function(df_sub, l, B){
    ##generate blocks
    b <- length(df_sub) - l + 1
    blocks <- matrix(NA, b, l)
    for (i in 1:b){blocks[i, ] <- df_sub[i:(i + l - 1)]}
    ##generate bootstrap counterparts
    bs <- round(length(df_sub) %/% l)
    X_mean <- rep(NA, B)
    for (i in 1:B){
      idx <- sample(b, bs, replace = T)
      tmp <- c()
      for (j in idx){tmp <- c(tmp, blocks[j, ])}
      X_mean[i] <- mean(tmp)}
    ##generate phi_i
    phi_i <- mean(X_mean) - mean(df)
    ##output
    return (phi_i)
  }
  
  generate_mse <- function(l, B, m){
    phi_i_list <- rep(NA, n - m + 1)
    for (i in 1:(n - m + 1)){phi_i_list[i] <- generate_phi_i(subsets[i, ], l, B)}
    phi_l0 <- generate_phi_i(df, l0, B)
    mse <- mean((phi_i_list - phi_l0) ^ 2)
    return (mse)
  }
  
  ###MAIN###
  mse_list <- rep(NA, length(l))
  for (i in l){mse_list[i] <- generate_mse(i, B, m)}
  l_opt_m <- which.min(mse_list)
  l_opt <- l_opt_m * (n / m) ^ (1 / 3)
  
  ###OUTPUT###
  p <- ggplot() + geom_point(aes(x = l, y = mse_list), size = 2) +
    labs(title = 'Scatter Plot of MSE')+ theme_light()
  print(p)
  return (round(l_opt))
}
```

Choose $m=0.25n \approx 25,l_0=0.05n \approx 5$.
```{r, out.width='70%', fig.align='center', fig.width=6, fig.height=4}
set.seed(5201314)
l_opt <- Subsample_Bootstrap(data, 25, 5, 2000, 1:10)
```

View $l_{\text {opt}}$.
```{r}
print(paste('Optimal l =', l_opt))
```

## (b)

Obtain the observed data estimate for the 90th percentile of the annual change.
```{r}
x_90 <- quantile(data, 0.9)
as.numeric(x_90)
```

\newpage

Define the function of `Moving Block Bootstrap` method.
```{r}
Moving_Block <- function(df, l, B){
  
  ###INITAL VALUES###
  b <- length(df) - l + 1
  ##generate blocks
  blocks <- matrix(NA, b, l)
  for (i in 1:b){blocks[i, ] <- df[i:(i + l - 1)]}
  bs <- round(length(df) %/% l)
  X_90 <- rep(NA, B)

  ###MAIN###
  for (i in 1:B){
    idx <- sample(b, bs, replace = T)
    tmp <- c()
    for (j in idx){tmp <- c(tmp, blocks[j, ])}
    X_90[i] <- quantile(tmp, 0.9)}

  ###OUTPUT###
  return (X_90)
}
```

Bootstrap for 10000 times and plot the distribution of bootstrap estimates.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result <- Moving_Block(data, 5, 10000)
hist(result, main = 'Bootstrap Estimate of 90th Percentile of Annual Change')
```

Estimate the standard error of this estimate using the moving block bootstrap.
```{r}
print(paste('Standard Error:', sd(result)))
```

## (c)

### 1. Nonparametric Bootstrap

Bootstrap the residuals without any assumption on $\epsilon_t$.
$$
X_t=\alpha X_{t-1}+\epsilon_t
$$

Define the function of nonparametric `Model-Based` approach.
```{r}
Model_Based_np <- function(y, x, B){
  
  ###INITIAL VALUES###
  fit <- lm(y ~ x - 1)
  coef <- fit$coefficients
  res <- residuals(fit)
  X_90 <- rep(NA, B)
  
  ###MAIN###
  for (i in 1:B){
    res_bootstrap <- sample(res, length(res) + 2, replace = T)
    tmp <- res_bootstrap[1]
    for (j in 1:(length(res) + 1)){
      tmp <- c(tmp, coef %*% tmp[j] + res_bootstrap[j + 1])
    }
    X_90[i] <- quantile(tmp[-1], 0.9)
  }
  
  ###OUTPUT###
  hist(X_90, main = paste0('Bootstrap Estimate of 90th Percentile (', B, ')'))
  return (X_90)
}
```

\newpage

Histogram the bootstrap distribution.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result <- Model_Based_np(data[2:98], data[1:97], 10000)
```

Estimate the standard error for the estimate from part (b).
```{r}
print(paste('Standard Error:', sd(result)))
```

### 2. Parametric Bootstrap

Bootstrap the errors with an assumption of $\epsilon_t \sim N(0,\sigma^2)$.
$$
X_t=\alpha X_{t-1}+\epsilon_t, \epsilon_t \sim N(0,\sigma^2)
$$

Define the function of parametric `Model-Based` approach.
```{r}
Model_Based_p <- function(y, x, B){

  ###INITIAL VALUES###
  fit <- lm(y ~ x - 1)
  coef <- fit$coefficients
  res <- residuals(fit)
  X_90 <- rep(NA, B)
  
  ###MAIN###
  for (i in 1:B){
    res_bootstrap <- rnorm(length(res) + 2, 0, sd(res))
    tmp <- res_bootstrap[1]
    for (j in 1:(length(res) + 1)){
      tmp <- c(tmp, coef %*% tmp[j] + res_bootstrap[j + 1])
    }
    X_90[i] <- quantile(tmp[-1], 0.9)
  }
  
  ###OUTPUT###
  hist(X_90, main = paste0('Bootstrap Estimate of 90th Percentile (', B, ')'))
  return (X_90)
}
```

Histogram the bootstrap distribution.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result <- Model_Based_p(data[2:98], data[1:97], 10000)
```

Estimate the standard error for the estimate from part (b).
```{r}
print(paste('Standard Error:', sd(result)))
```

So we may conclude:

- Standard Error: Nonparametric Model-based Bootstrap > Moving Block Bootstrap > Parametric Model-based Bootstrap.

\newpage

## (d)

Obtain the observed data estimate for the lag-1 autocorrelation of the annual change.
$$
\hat{r}=\sum_{t=1}^{n-1}\left(X_{t}-M\right)\left(X_{t+1}-M\right) / \sum_{t=1}^{n}\left(X_{t}-M\right)^{2}
$$

Where $n$ is the length of the series and $M$ is the mean of $X_1, ..., X_n$.

```{r}
m <- mean(data)
rho_hat <- sum((data[1:97] - m) * (data[2:98] - m)) / sum((data - m) ^ 2); rho_hat
```

$\mathcal{X}$ yields 97 couples $Y_t=(X_t,X_{t+1})$, and the vectorized series is $\mathcal{Y}=(Y_1, Y_2, ... , Y_{97})$. From these 97 blocks, we may resample blocks of blocks. Let each of these blocks of blocks be comprised of 5 of the small blocks. The lag 1 autocorrelation can be estimated as:
$$
\hat{r}^*=\sum_{t=1}^{n^*}\left(Y^*_{t, 1}-M\right)\left(Y^*_{t, 2}-M\right) / \sum_{t=1}^{n}\left(X_{t}-M\right)^{2}
$$

Where $Y^*_{t,j}$ is the $j^{th}$ element in $Y^*_t$ and $n^*$ is the length of $Y^*_t$.

Define the function of the moving block bootstrap with `Blocks-of-Blocks` strategy.
```{r}
Blocks_of_Blocks <- function(df, l, B, method = NA){
  
  ###INITIAL VALUES###
  rho_list <- rep(NA, B)
  M <- mean(df)
  b <- length(df) - 1
  ##generate blocks of blocks
  smallblocks <- rbind(df[1:b], df[2:(b + 1)])
  bigblocks <- array(NA, c(2, l, b - l + 1))
  for (i in 1:(b - l + 1)){bigblocks[, , i] <- smallblocks[, i:(i + l - 1)]}
  bs <- round(length(df) %/% l)
  
  ###FUNCTIONS###
  calculate_rho <- function(y){
    if (is.na(method)){sum((y[1, ] - M) * (y[2, ] - M)) / sum((df - M) ^ 2)}
    else {cor(y[1, ], y[2, ])}
  }
  
  ###MAIN###
  for (i in 1:B){
    idx <- sample(1:(b - l + 1), bs, replace = T)
    tmp <- c(); for (j in idx){tmp <- cbind(tmp, bigblocks[, , j])}
    rho_list[i] <- calculate_rho(tmp)
  }
  
  ###OUTPUT###
  hist(rho_list, main = paste0('Blocks-of-Blocks Bootstrap (', B, ')'))
  return(rho_list)
}
```

Histogram the bootstrap distribution of lag-1 autocorrelation of the annual change.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result <- Blocks_of_Blocks(data, 5, 10000)
```

Obtain the standard error of the estimate.
```{r}
print(paste('Standard Error:', sd(result)))
```

Obtain the bootstrap bias.
$$
\hat \Delta=\frac {1}{B}\sum_{i=1}^{B}(\hat r^*-\hat r)=\frac {1}{B}\sum_{i=1}^{B}\hat r^*-\hat r
$$

```{r}
mean(result) - rho_hat
```

We can also obtain the observed data estimate with `cor` function.
```{r}
rho_hat <- cor(data[1:97], data[2:98]); rho_hat
```

Using this method to estimate lag-1 autocorrelation and bootstrap the distribution.
```{r, out.width='70%', fig.align='center'}
set.seed(5201314)
result <- Blocks_of_Blocks(data, 5, 10000, 'cor')
```

Obtain the standard error of the estimate.
```{r}
print(paste('Standard Error:', sd(result)))
```

Obtain the bootstrap bias.
```{r}
mean(result) - rho_hat
```

So we can conclude that the first estimator is inferior to the second estimator because the former is more biased (due to approximation) and less efficient.
